---
title: "01-molecular-subtype-pineoblastoma"
author: "Zhuangzhuang Geng"
date: "2023-12-06"
output: html_document
---

## load library
```{r}
library(tidyverse)

```

## set directories
```{r}

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "molecular-subtyping-PB")
result_dir <- file.path(analysis_dir, "results")
if(!dir.exists(result_dir)){
  dir.create(result_dir)
}

```

## read files
```{r}

histo <- readr::read_tsv(file.path(data_dir, "histologies-base.tsv"))

## json file
path_dx_list_pb <- jsonlite::fromJSON(
  file.path(analysis_dir, "PB-subset",  
            "PB_subtyping_path_dx_strings.json")
)
```

## select PB samples from histologies
```{r}
pb_meta <- histo %>% 
  filter(pathology_diagnosis %in% path_dx_list_pb$exact_path_dx | 
           pathology_free_text_diagnosis %in% path_dx_list_pb$path_free_text_exact) %>% 
  mutate(molecular_subtype_methyl = case_when(
    dkfz_v12_methylation_subclass == "PB_FOXR2" ~ "PB, MYC/FOXR2", 
    dkfz_v12_methylation_subclass == "PIN_RB" ~ "PB, RB1",
    dkfz_v12_methylation_subclass == "PB_GRP1A" ~ "PB, Group 1A",
    dkfz_v12_methylation_subclass == "PB_GRP1B" ~ "PB, Group 1B",
    dkfz_v12_methylation_subclass == "PB_GRP2" ~ "PB, Group 2")) 

```

## there are four subtypes: 
### pineoblastoma, miRNA processing-altered 1 (PB, Group 1)
### pineoblastoma, miRNA processing-altered 2 (PB, Group 2)
### pineoblastoma, RB1-altered (PB, RB1)
### pineoblastoma, MYC/FOXR2-activated (PB, MYC/FOXR2)

## for now, subtyping is based on the methylation classification 
## PB_FOXR2: pineoblastoma, MYC/FOXR2-activated (PB, MYC/FOXR2)
## PB_GRP1A, PB_GRP1B: pineoblastoma, miRNA processing-altered 1 (PB, Group 1)
## PB_GRP2: pineoblastoma, miRNA processing-altered 2 (PB, Group 2)
## PIN_RB: pineoblastoma, RB1-altered (PB, RB1)

```{r}

pb_methy_with_subtype <- pb_meta %>% 
  filter(experimental_strategy == "Methylation" & dkfz_v12_methylation_subclass_score >= 0.8) %>% 
  mutate(molecular_subtype = case_when(molecular_subtype_methyl == "PB, MYC/FOXR2" ~ "PB, MYC/FOXR2", 
                                       molecular_subtype_methyl == "PB, RB1" ~ "PB, RB1",
                                       molecular_subtype_methyl %in% c("PB, Group 1A", "PB, Group 1B") ~ "PB, Group 1",
                                       molecular_subtype_methyl == "PB, Group 2" ~ "PB, Group 2"))
  

pb_methy_without_subtype <- pb_meta %>% 
  filter(experimental_strategy == "Methylation" & 
           !(Kids_First_Biospecimen_ID %in% pb_methy_with_subtype$Kids_First_Biospecimen_ID))

pb_methyl_df <- pb_methy_with_subtype %>% 
  bind_rows(pb_methy_without_subtype) %>% 
  select(match_id, molecular_subtype, molecular_subtype_methyl)

```

## assign the molecular subtypes to all samples based on the methylation results
```{r}

pb_meta_with_mol <- pb_meta %>%
  select(-molecular_subtype_methyl, -molecular_subtype) %>%
  left_join(pb_methyl_df, by = "match_id") %>% 
  select(Kids_First_Biospecimen_ID, Kids_First_Participant_ID, match_id, sample_id, molecular_subtype_methyl, molecular_subtype) %>% 
  ## Assign samples without subtype as "PB, To be classified"
  mutate(molecular_subtype = case_when(is.na(molecular_subtype) ~ "PB, To be classified", 
                                       TRUE ~ molecular_subtype)) %>%
  arrange(match_id) %>%
  write_tsv(file.path(result_dir, "pineo-molecular-subtypes.tsv"))

# do all bs ids have unique subtypes?
length(unique(pb_meta_with_mol$Kids_First_Biospecimen_ID)) == nrow(pb_meta_with_mol)

```

## session info

```{r}

sessionInfo()
```



