---
title: "Subtype Neurocytoma tumors as central or extra-ventricular"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Krutika Gaonkar for D3b
date: 2020
---

For, "Neurocytoma" samples detected in pathology_diagnosis, subtyping conditions as per issue [#805](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/805) are as follows :

 - If primary_site == "Ventricles", then molecular_subtype == CNC (central neurocytoma) 
 - If primary_site != "Ventricles", then molecular_subtype == EVN (extraventricular neurocytoma)

## Set up 

```{r}
library(tidyverse)
```

## Input directories and files

```{r}
data_dir <- file.path("..", "..", "data")
histologies_df <- read_tsv(file.path(data_dir, "histologies-base.tsv"))
path_dx_list <-  jsonlite::fromJSON(
  file.path("neurocytoma-subset","neurocytoma_subtyping_path_dx_strings.json"))


```

## Output directory and file

```{r}
results_dir <- "results"
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
outputfile <- file.path(results_dir,"neurocytoma_subtyping.tsv")
```


## Subset neurocytoma samples from histologies based on the pathology_diagnosis json file.

```{r}

neurocytoma_from_pathology_diagnosis <- histologies_df %>%
  filter(pathology_diagnosis %in% path_dx_list$exact_path_dx) 


```
There are no Kids_First_Biospecimen_ID in `pathology_free_text_diagnosis` that are not listed as "Neurocytoma" in  `pathology_diagnosis`. So will only use  `pathology_diagnosis` to look up "Neurocytoma" samples for subtyping


Checking if `pathology_diagnosis` has "Neurocytoma" what are all primary_site associated with it
```{r}
neurocytoma_from_pathology_diagnosis$primary_site %>% unique()
```


## Subtyping Neurocytoma samples based on primary site 

The values are only "Ventricles", "Parietal Lobe", or "Frontal Lobe;Temporal Lobe" so we can directly check if primary_site== "Ventricles" ~ "CNC" and if primary_site != "Ventricles" ~ "EVN"

```{r}
# Get methylation subtypes
neurocytoma_subtype <- neurocytoma_from_pathology_diagnosis %>% 
  filter(experimental_strategy == "Methylation", 
         dkfz_v12_methylation_subclass_score >= 0.8) %>%
  mutate(methyl_subtype = case_when(dkfz_v12_methylation_subclass == "EVNCYT" ~ "EVN",
                                           dkfz_v12_methylation_subclass == "CN" ~ "CNC",
                                           TRUE ~ NA_character_)) %>%
  select(sample_id, composition, methyl_subtype, Kids_First_Biospecimen_ID) %>%
  unique()
  
# add methyl with low score (for future analysis)

methyl_no_subtype <- neurocytoma_from_pathology_diagnosis %>% 
  filter(experimental_strategy == "Methylation", 
         !Kids_First_Biospecimen_ID %in% neurocytoma_subtype$Kids_First_Biospecimen_ID) %>% 
  mutate(methyl_subtype = NA_character_) %>%
  select(sample_id, composition, methyl_subtype, Kids_First_Biospecimen_ID) %>%
  unique()

neurocytoma_subtype <- bind_rows(neurocytoma_subtype, methyl_no_subtype) %>% 
  select(sample_id, composition, methyl_subtype)

neurocytoma_subtype <- neurocytoma_from_pathology_diagnosis %>% 
  dplyr::mutate(molecular_subtype= 
                  # primary_site == "Ventricles", 
                  # then molecular_subtype == CNC (central neurocytoma) 
                  case_when(primary_site == "Ventricles" ~ "CNC",
                  # primary_site != "Ventricles", 
                  # then molecular_subtype == EVN (extraventricular neurocytoma)       
                            primary_site != "Ventricles" ~ "EVN")) %>%
  left_join(neurocytoma_subtype) %>%
  dplyr::select(Kids_First_Participant_ID, sample_id, composition, Kids_First_Biospecimen_ID,
                molecular_subtype, methyl_subtype) %>%
  arrange(sample_id) %>%
  write_tsv(outputfile)

```
