---
title: "Add molecular subtype for samples that have hallmark Ewings Sarcoma fusions"
output: html_notebook
author: K S Gaonkar for D3b
date: January 2020
---

Identify sample IDs with hallmark _EWSR1_ fusions and subtype as `EWS` 

```{r}
library("tidyverse")
```

### Set directories and file paths

```{r}
# to get root directory
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
dataDir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "molecular-subtyping-EWS")
results_dir <- file.path(analysis_dir, "results")

# create output file
ews_file <- file.path(results_dir, "EWS_results.tsv")
```

# Read in data
```{r read in data}
# histology file
clinical <- readr::read_tsv(file.path(dataDir, "histologies-base.tsv"), guess_max = 100000) %>%
  filter(cohort %in% c("PBTA", "Kentucky", "DGD"))

# read pathology diagnosis JSON file 
path_dx_list <- jsonlite::fromJSON(file.path(analysis_dir, "ews-subset", "EWS_subtyping_path_dx_strings.json"))

# ewings samples or samples with EWS via methylation
ews_clinical <- clinical %>%
  filter(pathology_diagnosis == path_dx_list$exact_path_dx | (dkfz_v12_methylation_subclass == "EWS" & dkfz_v12_methylation_subclass_score >= 0.8)) 

# what are the path diagnoses here? Non-EWS should be reviewed by pathology
ews_clinical %>%
  count(pathology_diagnosis) %>%
  arrange(desc(n))

```

# get ewings fusion calls from fusion-summary results file
```{r fusion bsids}
ewings_foi_bsids <- readr::read_tsv(
  file.path(root_dir, 
            "data", 
            "fusion_summary_ewings_foi.tsv")
  ) 

# get subset folder
results_dir <- "results"

# create if doesn't exist
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```

### Get sample_id from RNA-Seq samples that have ewings sarcoma EWSR1 fusion

```{r}
# get a vector of biospecimen IDs for those biospecimens that should be subtyped
has_ews_fusion <- rowSums(ewings_foi_bsids[, -1]) > 0

# get bs_ids which have hallmark fusion 
fusion_bsid <- ewings_foi_bsids %>%
  filter(has_ews_fusion) %>%
  left_join(clinical) %>%
  filter(cohort %in% c("PBTA", "Kentucky", "DGD")) %>%
  select(Kids_First_Biospecimen_ID, sample_id, composition) %>%
  mutate(molecular_subtype = "EWS",
         id = paste(sample_id, composition, sep = "_"))

# subtype map 
subtype_map <- fusion_bsid %>%
  select(id, molecular_subtype) %>%
  unique()

```

# get methylation EWS
```{r methylation subtypes}
# get all methyl samples
all_methyl <- clinical %>%
  filter(experimental_strategy == "Methylation") %>%
  pull(Kids_First_Biospecimen_ID)

# we will add methyl subtypes 
methyl_bs_ids_with_subtypes <- clinical %>%
  filter(experimental_strategy == "Methylation",
         dkfz_v12_methylation_subclass == "EWS" & dkfz_v12_methylation_subclass_score >= 0.8) %>%
  # reformat classifier result per https://www.molecularneuropathology.org/mnp/classifiers/11
  mutate(molecular_subtype_methyl = dkfz_v12_methylation_subclass) %>%
  mutate(id = paste(sample_id, composition, sep = "_")) %>%
  dplyr::select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, sample_id, composition, tumor_descriptor, molecular_subtype_methyl, id) 

# add samples with poor methyl score

methy_no_subtype <- clinical %>% 
  filter(experimental_strategy == "Methylation",
         dkfz_v12_methylation_subclass == "EWS" & dkfz_v12_methylation_subclass_score < 0.8) %>% 
  mutate(molecular_subtype_methyl = NA_character_) %>%
  mutate(id = paste(sample_id, composition, sep = "_")) %>%
  dplyr::select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, sample_id, composition, tumor_descriptor, molecular_subtype_methyl, id)

methyl_map <- bind_rows(methyl_bs_ids_with_subtypes, methy_no_subtype)

# methyl map
methyl_map <- methyl_map %>%
  select(id, molecular_subtype_methyl) %>%
  unique()


```

### Get other samples matching above ids

```{r}

all_subtypes <- ews_clinical %>%
  select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID, sample_id, composition, pathology_diagnosis) %>%
  mutate(id = paste(sample_id, composition, sep = "_")) %>%
  left_join(subtype_map) %>%
  left_join(methyl_map) %>%
  # confirm the subtype based in the pathology, remove any with different pathology dx
  filter(pathology_diagnosis == "Ewings Sarcoma") %>% 
  # make subtype EWS if diagnosis EWS or methylation was EWS
  mutate(molecular_subtype = case_when(pathology_diagnosis == "Ewings Sarcoma" ~ "EWS",
                                       TRUE ~ molecular_subtype_methyl))

```

#### To investigate!
BS_ER9RMSVG - this sample has MB diagnosis. It was To be classified via MB subtyping.
BS_SRW0DH5G - this sample has Cranio diagnosis. It was To be classified via Cranio subtyping.
These may be mis-diagnoses. Will need to confirm with pathology.

#### Save to file

```{r}
all_subtypes %>%
  readr::write_tsv(ews_file)
```

# get session info
```{r}
sessionInfo()
```



