---
title: "Subtyping chordoma"
output: html_notebook
author: Mateusz Koptyra 
date: 20191121
---

This notebook prepares _SMARCB1_ copy number and expression data for chordoma samples for the purpose of identifying poorly-differentiated chordoma samples, which are characterized by loss of _SMARCB1_.

## Set up

```{r}
library(dplyr)
library(readr)
library(ggplot2)
```

### Read in data

```{r}
# File path to subsetted files directory
input_dir <- "chordoma-subset"
```

Set up metadata

```{r}
# Read in non-subsetted metadata
histologies_df <- readr::read_tsv(file.path("..", "..", "data", "histologies-base.tsv"), guess_max = 100000)

# Read pathology diagnosis json file
path_dx_list <- jsonlite::fromJSON(
  file.path(input_dir,
            "chordoma_subtyping_path_dx_strings.json")
)

# Subset metadata
subset_metadata <- histologies_df %>%
  dplyr::filter(pathology_diagnosis == path_dx_list$exact_path_dx) %>%
  select(Kids_First_Biospecimen_ID, sample_id, composition, Kids_First_Participant_ID,experimental_strategy,tumor_ploidy) %>%
  dplyr::rename(ploidy=tumor_ploidy) %>%
  mutate(id = paste(sample_id, composition, sep = "_"))

# collect methyl and Targeted Seq samples to add back later (these don't have CN or RNA expression)
methyl_panel <- subset_metadata %>%
  filter(experimental_strategy %in% c("Methylation", "Targeted Sequencing")) %>%
  select(Kids_First_Biospecimen_ID, sample_id, composition, Kids_First_Participant_ID,experimental_strategy,ploidy, id)
```

```{r}
subset_focal_cn_df <-
  data.table::fread(file.path(input_dir, "chordoma-only_cn_autosomes.tsv.gz"))
```

```{r}
# we need to include the sample_id field from histologies-base.tsv in the final
# table (field will allow #us to map between RNA-seq (e.g., SMARCB1 expression
# values) and WGS data (e.g., SMARCB1 focal copy number status) from the same
# event for a given individual). To get the SMARCB1 jitter plot in the photo
# here #250 (comment), you will first need to read in the collapsed expression
# data
subset_expression_data <-
  read_rds(file.path(
    input_dir,
    "chordoma-only-gene-expression-rsem-tpm-collapsed.rds"
  ))
```

### Output

```{r}
# tabular data
results_dir <- "results"
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
output_file <- file.path(results_dir, "chordoma_smarcb1_status.tsv")
```

## Prepare the data

Extracting the chordoma samples that have a loss of SMARCB1 from `subset_focal_cn_df`

```{r}
chordoma_loss <- subset_focal_cn_df %>%
  filter(
    gene_symbol == "SMARCB1",
    status == "loss"
  )
chordoma_loss
```

Distinguishing the chordoma samples with no copy number change chromosome 22
```{r}
copy_neutral_df <- subset_metadata %>%
  # the copy events can only be taken from WGS/WXS data not RNA-seq data
  # we also only want biospecimens where a loss was not recorded to avoid duplicates
  filter(
    experimental_strategy %in% c("WGS", "WXS"),
    !(Kids_First_Biospecimen_ID %in% chordoma_loss$biospecimen_id)
  ) %>%
  # if there's no loss, let's assume status is copy neutral
  mutate(status = "neutral",
         copy_number = ploidy) %>%
  select(Kids_First_Biospecimen_ID, id, status, copy_number,ploidy)
copy_neutral_df
```

Joining chordoma samples (both with chr 22 loss and no copy number change)
```{r}
chordoma_copy <- chordoma_loss %>%
  dplyr::rename(Kids_First_Biospecimen_ID = biospecimen_id) %>%
  # join the losses with the neutrals to get a new data frame
  select(Kids_First_Biospecimen_ID, status,ploidy,copy_number) %>%
  left_join(subset_metadata[,c("Kids_First_Biospecimen_ID", "id")]) %>%
  bind_rows(copy_neutral_df)
chordoma_copy
```

Look at SMARCB1 expression values only in chordoma

```{r}
# get the row that contains the SMARCB1 values
# gene symbols are rownames
smarcb1_expression <- subset_expression_data[which(rownames(subset_expression_data) == "SMARCB1"), ]
```

```{r}
# now only the columns correspond to chordoma samples
smarcb1_expression <- smarcb1_expression[, which(colnames(subset_expression_data) %in% subset_metadata$Kids_First_Biospecimen_ID) ]

# Print out the expression for SMARCB1
smarcb1_expression
```

The `smarcb1_expression` is a not a friendly form ^^; Transposing needed: 

```{r}
# transpose such that samples are rows
smarcb1_expression <- t(smarcb1_expression) %>%
  # make a data.frame
  as.data.frame() %>%
  # we want the rownames that are biospecimen identifers as their own column called Kids_First_Biospecimen_ID
  tibble::rownames_to_column("Kids_First_Biospecimen_ID") %>%
  # give SMARCB1 column a slightly better column name
  dplyr::rename(SMARCB1_expression = SMARCB1) %>%
  left_join(subset_metadata[,c("Kids_First_Biospecimen_ID", "id")])
smarcb1_expression
```


Joining the copy number data with the expression data in this step
```{r}
chordoma_smarcb1_df <- smarcb1_expression %>%
  bind_rows(chordoma_copy) %>%
  # add pt id
  left_join(subset_metadata[,c("Kids_First_Participant_ID", "Kids_First_Biospecimen_ID")]) %>%
  # 'status' is replaced a more descriptive name
  dplyr::rename(focal_SMARCB1_status = status,
                SMARCB1_copy_number = copy_number) %>%
  select(
    Kids_First_Participant_ID,
    Kids_First_Biospecimen_ID,
    id,
    focal_SMARCB1_status,
    ploidy,
    SMARCB1_copy_number,
    SMARCB1_expression
  ) %>%
  # add remaining samples
  bind_rows(methyl_panel) %>%
  arrange(id)
chordoma_smarcb1_df
```


Write the table to file.

```{r}
chordoma_smarcb1_df %>%
  write_tsv(output_file)
```

### Session Info

```{r}
sessionInfo()
```
