---
title: "Annotate SNV subtype status for LGAT biospecimens"
output: html_notebook
author: K S Gaonkar 
date: 2020
---

In this PR we will use identify LGAT biospecimens from pathology diagnosis and annotate subtype specific SNV status per biospecimen.

As per [issue](https://github.com/AlexsLemonade/OpenPBTA-analysis/issues/790) we will be subtyping LGAT based on SNV in the following genes:

- LGG, NF1
somatic loss of NF1 via either missense, nonsense mutation 

- LGG, BRAF V600E
contains BRAF V600E or V599 SNV or non-canonical BRAF alterations such as p.V600ins or p.D594N

- LGG, other MAPK
contains KRAS, NRAS, HRAS, MAP2K1, MAP2K2, MAP2K1, ARAF SNV or indel

- LGG, RTK
harbors a MET SNV 
harbors a KIT SNV or
harbors a PDGFRA SNV 

- LGG, FGFR
harbors FGFR1 p.N546K, p.K656E, p.N577, or p. K687 hotspot mutations or

- LGG, IDH
harbors an IDH R132 mutation

- LGG, H3.3
harbors an H3-3A or H3-3B K28M or G35R/V mutation

-  LGG, H3.1
harbors an H3C2 K28M
harbors and H3C3  K28M


### Setup
```{r}
library(tidyverse)

# Look for git root folder
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

# Get subset folder
subset_dir <- file.path(root_dir, 
                        "analyses",
                        "molecular-subtyping-LGAT",
                        "lgat-subset")

# create if doesn't exist
if (!dir.exists(subset_dir)) {
  dir.create(subset_dir)
}
```

## Input

```{r}
# File from 00-LGAT-select-pathology-dx that is used for the pathology diagnosis
# inclusion/exclusion criteria
path_dx_list <- jsonlite::fromJSON(
  file.path(subset_dir, 
            "lgat_subtyping_path_dx_strings.json")
)

# clinical file
clinical <- read_tsv(file.path(root_dir, 
                               "data",
                               "histologies-base.tsv"), 
                     guess_max = 100000)
```

### SNV for subtyping 
 - We will use the `snv-consensus-plus-hotspots.maf.tsv.gz` which contained consensus calling plus hotspot rescued for SNV calling - detailed in [here](https://github.com/AlexsLemonade/OpenPBTA-analysis/tree/master/analyses/snv-callers#snv-caller-comparison-analysis)). Also include the DGD MAF, `snv-dgd.maf.tsv.gz` for OpenPedCan.

```{r load-inputs-please-wait}

# select tumor sample barcode, gene, short protein annotation, domains, and variant classification
keep_cols <- c("Chromosome",
             "Start_Position",
             "End_Position",
             "Strand",
             "Variant_Classification",
             "IMPACT",
             "Tumor_Sample_Barcode",
             "Hugo_Symbol",
             "HGVSp_Short",
             "SIFT",
             "PolyPhen",
             "DOMAINS")

#snv_dgd_maf <- data.table::fread(
 # file.path(root_dir, "data" , "snv-dgd.maf.tsv.gz"),
  #select = keep_cols,
  #data.table = FALSE) 

snv_consensus_hotspot_maf <- data.table::fread(
  file.path(root_dir, "data" , "snv-consensus-plus-hotspots.maf.tsv.gz"),
  select = keep_cols,
  data.table = FALSE) #%>% 
  #dplyr::bind_rows(snv_dgd_maf)

```

Filter to tumor samples that should be included on the basis of pathology diagnosis
```{r}
lgat_specimens_df <- clinical %>%
  dplyr::filter(str_detect(str_to_lower(pathology_diagnosis),  # Inclusion criteria
                           paste(path_dx_list$include_path_dx, collapse = "|")),
                # Exclusion criteria for pathology diagnosis
                str_detect(str_to_lower(pathology_diagnosis),
                           paste(path_dx_list$exclude_path_dx, collapse = "|"),
                           negate = TRUE),
                # Tumors
                sample_type == "Tumor",
                experimental_strategy != "Methylation")

# Write this intermediate file to the subset directory as it allows for
# inspection
write_tsv(lgat_specimens_df, file.path(subset_dir, "lgat_metadata.tsv"))

# Filter to dna samples
lgat_dna_df <- lgat_specimens_df %>%
  dplyr::filter(experimental_strategy == "WGS" | 
                  experimental_strategy == "WXS" | 
                  experimental_strategy == "Targeted Sequencing", 
                is.na(RNA_library)) %>%
  # will keep Kids_First_Biospecimen_ID 
  # sample_id is kept to be able to match with RNA-Seq in the later step
  dplyr::select(Kids_First_Biospecimen_ID, sample_id)
```


Gather gene(s) that define LGAT subtypes
Additional information for genes with known hotspots/canonical mutation is also provided in the list. 
If the gene has multiple hotspots we will gather the protein hotspots site to use it in the below chunks to grep the `HGVSp_Short` column.

```{r}
# combined list for SNV of interest per subtype
#
snvOI <- jsonlite::fromJSON(file.path(root_dir,
                                      "analyses",
                                      "molecular-subtyping-LGAT",
                                      "input",
                                      "snvOI_list.json"))

# Collapse multiple hotspots in genes with "|" so easy grep calls
BRAF_hotspot <-paste(snvOI$BRAF_V600E$hotspot[!is.na( snvOI$BRAF_V600E$hotspot)],collapse = "|")
FGFR_hotspot <-paste(snvOI$FGFR$hotspot[!is.na( snvOI$FGFR$hotspot)],collapse = "|")
IDH1_hotspot<- paste(snvOI$IDH1$hotspot[!is.na( snvOI$IDH1$hotspot)],collapse = "|")
```


### Subset consensus maf and annotate per subtype based on SNV

We will remove synonymous variant calls from the list above. The Variant_Classification terms for synonymous SNV are selected as per [interaction-plots/scripts/02-process_mutations.R](https://github.com/AlexsLemonade/OpenPBTA-analysis/blob/master/analyses/interaction-plots/scripts/02-process_mutations.R).
```{r}
# Variant Classification with Low/Modifier variant consequences 
#  from maftools http://asia.ensembl.org/Help/Glossary?id=535
synonymous <- c(
  "Silent",
  "Start_Codon_Ins",
  "Start_Codon_SNP",
  "Stop_Codon_Del",
  "De_novo_Start_InFrame",
  "De_novo_Start_OutOfFrame"
)

snv_consensus_hotspot_maf <- snv_consensus_hotspot_maf %>%
  dplyr::filter(!Variant_Classification %in% synonymous)
```

 We will also remove the LOW/MODIFIER IMPACT annotated by [VEP](https://docs.gdc.cancer.gov/Data/File_Formats/MAF_Format/#impact-categories)
 
 - LOW (L): Assumed to be mostly harmless or unlikely to change protein behavior
 - MODIFIER (MO): Usually non-coding variants or variants affecting non-coding genes, where predictions are difficult or there is no evidence of impact

```{r}
snv_consensus_hotspot_maf <- snv_consensus_hotspot_maf %>%
  dplyr::filter(!IMPACT %in% c("MODIFIER","LOW"))
```

### Gather status of all SNVs per LGAT biospecimen

We will gather SNV calls that satisfy the conditions per subtype and save the subset along with the HGVSp_Short, DOMAINS and Variant_Classification 
and recode them as subtypes
```{r}
# Filter consensus mutation files for LGAT subset
snv_consensus_hotspot_maf <- snv_consensus_hotspot_maf %>%
  # find lgat samples
  dplyr::filter(Tumor_Sample_Barcode %in% lgat_dna_df$Kids_First_Biospecimen_ID)


# Get BRAF mutation status
# canonical mutations V600E
lgat_braf_V600 <- snv_consensus_hotspot_maf %>%
  dplyr::filter(
    HGVSp_Short %in% snvOI$BRAF_V600E$canonical[!is.na(snvOI$BRAF_V600E$canonical)] &
      Hugo_Symbol=="BRAF" | # OR
      # hotspot mutations in p.600 and p.599
      grepl(BRAF_hotspot,HGVSp_Short) &
      Hugo_Symbol=="BRAF" | # OR
      # and kinase domain mutation for non-canonical mutation 
      # Family: PK_Tyr_Ser-Thr https://pfam.xfam.org/family/PF07714
      grepl("PF07714",DOMAINS) & 
      Hugo_Symbol=="BRAF") %>%
  right_join(lgat_dna_df,by=c("Tumor_Sample_Barcode"="Kids_First_Biospecimen_ID"))

# Get other MAPK mutation status
# all mutations in MAPK genes
lgat_mapk <- snv_consensus_hotspot_maf %>%
  dplyr::filter(
    Hugo_Symbol %in% snvOI$MAPK$gene) %>%
    # remove BRAF mutations in canonical V600E region
    dplyr::filter(!(Hugo_Symbol=="BRAF" &
                      HGVSp_Short %in% lgat_braf_V600$HGVSp_Short)) %>%
  right_join(lgat_dna_df,by=c("Tumor_Sample_Barcode"="Kids_First_Biospecimen_ID"))

# Get NF1 mutation status
lgat_nf1 <- snv_consensus_hotspot_maf %>%
  dplyr::filter(
    Hugo_Symbol %in% snvOI$NF1$gene & 
      Variant_Classification %in% c("Missense_Mutation","Nonsense_Mutation", "Splice_Site", "Translation_Start_Site", "Frame_Shift_Del", "Frame_Shift_Ins", "Splice_Region") 
  ) %>%
  right_join(lgat_dna_df,by=c("Tumor_Sample_Barcode"="Kids_First_Biospecimen_ID"))

# Get RTK mutation status
# all mutations in RTK genes
lgat_rtk <- snv_consensus_hotspot_maf %>%
  dplyr::filter(
    Hugo_Symbol %in% snvOI$RTK$gene 
  ) %>%
  right_join(lgat_dna_df,by=c("Tumor_Sample_Barcode"="Kids_First_Biospecimen_ID")) 

# Get FGFR mutation status
# canonical mutations     
lgat_fgfr <- snv_consensus_hotspot_maf %>%
  dplyr::filter(
    HGVSp_Short %in% snvOI$FGFR$canonical[!is.na(snvOI$FGFR$canonical)] &
      Hugo_Symbol=="FGFR1" | # OR
      # hotspot mutations 
      grepl(FGFR_hotspot,HGVSp_Short) &
      Hugo_Symbol=="FGFR1"
  ) %>%
  right_join(lgat_dna_df,by=c("Tumor_Sample_Barcode"="Kids_First_Biospecimen_ID"))

# Get IDH mutation status
# hostspot mutations
lgat_idh <- snv_consensus_hotspot_maf %>%
  dplyr::filter(
    grepl(IDH1_hotspot,HGVSp_Short) & 
      Hugo_Symbol %in% snvOI$IDH1$gene 
  ) %>%
  right_join(lgat_dna_df,by=c("Tumor_Sample_Barcode"="Kids_First_Biospecimen_ID"))

# Get histone mutation status
# H3F3A canonical mutations
lgat_histone <- snv_consensus_hotspot_maf  %>%
  dplyr::filter(
    HGVSp_Short %in% snvOI$`H3-3A`$canonical & Hugo_Symbol %in% "H3-3A" | # OR
      # H3F3B canonical mutations
      HGVSp_Short %in% snvOI$`H3-3B`$canonical & Hugo_Symbol %in% "H3-3B" | # OR
      # HIST1H3B canonical mutations
      HGVSp_Short %in% snvOI$H3C2$canonical & Hugo_Symbol %in% "H3C2" | # OR
      # HIST1H3C canonical mutations
      HGVSp_Short %in% snvOI$H3C3$canonical & Hugo_Symbol %in% "H3C3" | # OR
      # HIST2H3C canonical mutations     
      HGVSp_Short %in% snvOI$H3C14C$canonical & Hugo_Symbol %in% "H3C14" 
  ) %>%
  right_join(lgat_dna_df,by=c("Tumor_Sample_Barcode"="Kids_First_Biospecimen_ID"))

```


### Recode mutations
To gather the biospecimens, we will use all LGAT DNA biospecimen IDs and assign "Yes" or "No" to each subtype column depending on the presence or absence of SNV that define the subtype.

columnname  | description 
 --- | --- 
NF1_mut | somatic loss of NF1 via either missense, nonsense mutation 
BRAF_V600E_mut | contains BRAF V600E or V599 SNV or non-canonical BRAF alterations such as p.V600ins or p.D594N 
MAPK_mut | contains mutation in KRAS, NRAS, HRAS, MAP2K1, MAP2K2, MAP2K1, ARAF SNV or indel 
RTK_mut | harbors a MET, KIT, or PDGFRA SNV 
FGFR_mut | harbors FGFR1 p.N546K, p.K656E, p.N577, or p. K687 hotspot mutations 
IDH_mut | harbors an IDH R132 mutation 
H3.1_mut | harbors an H3C2 K28M or H3C3  K28M
H3.2_mut | harbors an H3C14 K28M
H3.3_mut | harbors an H3-3A K28M or G35R/V mutation 

```{r}

recode_mutations <- function(.data, genes, subtype) {
  .data %>%
  reshape2::dcast(Tumor_Sample_Barcode + sample_id~ Hugo_Symbol,
                  fun.aggregate = length) %>%
  dplyr::mutate(!!as.name(subtype) := if_else(any(colnames(.) %in% genes) & rowSums(dplyr::select(.,one_of(genes))) >= 1,
          "Yes", 
          "No")
          )
}

lgat_subtypes_list <- list(recode_mutations(lgat_braf_V600,genes = snvOI$BRAF_V600E$gene, subtype = "BRAF_V600E_mut"),
                           recode_mutations(lgat_fgfr,genes = snvOI$FGFR$gene, subtype = "FGFR_mut"),
                           recode_mutations(lgat_idh,genes = snvOI$IDH1$gene,subtype = "IDH_mut"),
                           recode_mutations(lgat_histone,genes = c("H3C2","H3C3"),subtype = "H3.1_mut"),
                           recode_mutations(lgat_histone,genes = snvOI$H3C14$gene,subtype = "H3.2_mut"),
                           recode_mutations(lgat_histone,genes = c("H3-3A","H3-3B") ,subtype  = "H3.3_mut"),
                           recode_mutations(lgat_mapk,genes = snvOI$MAPK$gene, subtype = "MAPK_mut"),
                           recode_mutations(lgat_rtk,genes = snvOI$RTK$gene, subtype = "RTK_mut"),
                           recode_mutations(lgat_nf1,genes = snvOI$NF1$gene, subtype = "NF1_mut"))    

lgat_subtypes <- lgat_subtypes_list %>%
  reduce(inner_join, by = c("Tumor_Sample_Barcode","sample_id")) %>%  
  dplyr::rename("Kids_First_Biospecimen_ID"="Tumor_Sample_Barcode")  %>%
  dplyr::select("Kids_First_Biospecimen_ID","sample_id",ends_with("mut")) %>%
  dplyr::arrange(Kids_First_Biospecimen_ID, sample_id)
lgat_subtypes

```

Save to subset folder

```{r}

write_tsv(lgat_subtypes,file.path(subset_dir, "LGAT_snv_subset.tsv"))

```

### Methylation classification subtypes

```{r}
# high confidence methylation classification subtypes
lgat_methyl_subtyped <- clinical %>% 
  dplyr::filter(str_detect(str_to_lower(pathology_diagnosis), 
                           paste(c(path_dx_list$include_path_dx, "Dysplasia/Gliosis"), collapse = "|")),
                # Exclusion criteria for pathology diagnosis
                str_detect(str_to_lower(pathology_diagnosis),
                           paste(path_dx_list$exclude_path_dx, collapse = "|"),
                           negate = TRUE),
                dkfz_v12_methylation_subclass_score >= 0.8,
                experimental_strategy == "Methylation", 
                sample_type == "Tumor", 
                cohort != "TARGET") %>% 
  dplyr::mutate(molecular_subtype_methyl = 
                  case_when(grepl("LGG_MYB", dkfz_v12_methylation_subclass) ~ "LGG, MYB/MYBL1",
                            dkfz_v12_methylation_subclass == "SEGA" ~ "SEGA, To be classified",
                            dkfz_v12_methylation_subclass == "A_IDH_LG" ~ "LGG, IDH",
                            dkfz_v12_methylation_subclass == "PA_INF_FGFR"  ~ "LGG, FGFR",
                            dkfz_v12_methylation_subclass == "PA_INF"  ~ "LGG, BRAF/MAPK",
			                      dkfz_v12_methylation_subclass %in% c("PA_MID", "PLNTY")  ~ "LGG, MAPK",
                            dkfz_v12_methylation_subclass == "AG_MYB" ~ "LGG, MYB/MYBL1",
                            dkfz_v12_methylation_subclass == "GG" ~ "GNG, MAPK",
                  			    dkfz_v12_methylation_subclass == "RGNT" ~ "GNT, FGFR",
                  			    dkfz_v12_methylation_subclass %in% c("DLGNT_1", "DLGNT_2", "PA_CORT") ~ "GNT, MAPK",
                  			    dkfz_v12_methylation_subclass %in% c("DGONC", "DIG_DIA", "DNET", "GNT_A") ~ "GNT",
                  			    dkfz_v12_methylation_subclass == "PXA" ~ "PXA")) %>% 
  dplyr::select(Kids_First_Biospecimen_ID, Kids_First_Participant_ID, sample_id, molecular_subtype_methyl) %>%
  dplyr::rename("Kids_First_Biospecimen_ID_Methyl" = "Kids_First_Biospecimen_ID") %>% 
  dplyr::filter(!is.na(molecular_subtype_methyl)) %>% 
  dplyr::distinct()

# low confidence methylation classification subtypes
lgat_methyl_not_subtyped <- clinical %>% 
  dplyr::filter(str_detect(str_to_lower(pathology_diagnosis), 
                           paste(c(path_dx_list$include_path_dx, "Dysplasia/Gliosis"), collapse = "|")),
                # Exclusion criteria for pathology diagnosis
                str_detect(str_to_lower(pathology_diagnosis),
                           paste(path_dx_list$exclude_path_dx, collapse = "|"),
                           negate = TRUE),
                !Kids_First_Biospecimen_ID %in% unique(lgat_methyl_subtyped$Kids_First_Biospecimen_ID_Methyl),
                experimental_strategy == "Methylation", 
                sample_type == "Tumor", 
                composition == "Solid Tissue",
                cohort != "TARGET") %>%
  dplyr::left_join(lgat_methyl_subtyped %>% dplyr::select(Kids_First_Participant_ID, sample_id, molecular_subtype_methyl),
                   by = c("Kids_First_Participant_ID", "sample_id")) %>%
  dplyr::select(Kids_First_Biospecimen_ID, Kids_First_Participant_ID, sample_id, molecular_subtype_methyl) %>%
  dplyr::rename("Kids_First_Biospecimen_ID_Methyl" = "Kids_First_Biospecimen_ID") %>% 
  dplyr::distinct()

# merge methyl 
lgat_methyl_subtypes = dplyr::bind_rows(lgat_methyl_subtyped, lgat_methyl_not_subtyped)  
```

Save subtypes to folder

```{r}


readr::write_tsv(lgat_methyl_subtypes, file.path(subset_dir, "LGAT_methyl_subtypes.tsv"))

```

## Session Info

```{r}
sessionInfo()
```
