---
title: "DBS Mutational Signatures Analysis"
output: 
  html_notebook:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: 2022
---
  
  **Purpose:**
  
Calculate and plot DBS mutational signatures for all samples using [COSMIC signatures](https://cancer.sanger.ac.uk/cosmic).

#### Usage

To run this from the command line, use:
```
Rscript -e "rmarkdown::render('analyses/mutational-signatures/02-dbs_mutational_signatures.Rmd', 
                              clean = TRUE)"
```

_This assumes you are in the top directory of the repository._

## Setup

#### Packages and functions

Import necessary functions.

```{r load libraries}
# Magrittr pipe
`%>%` <- dplyr::`%>%`

# Import specialized functions
source(file.path("util", "mut_sig_functions.R"))

# Load this library
library(deconstructSigs)
library(tidyverse)
```

Set up directory paths. 

```{r set directories}
data_dir <- file.path("..", "..", "data")
input_dir <- "input"
results_dir <- "results"
plots_dir <- "plots"
figures_dir <- file.path("..", "..", "figures")
cosmic_dbs_plots <- file.path(plots_dir, "cosmicv3_dbs")

scratch_dir <- file.path("..", "..", "scratch", "mutational-signatures")
cosmicv3_dbs_scratch <- file.path(scratch_dir, "cosmicv3_dbs")
```

Make new directories for the results. 

```{r make directories}
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
if (!dir.exists(cosmic_dbs_plots)) {
  dir.create(cosmic_dbs_plots, recursive = TRUE)
}

if (!dir.exists(scratch_dir)) {
  dir.create(scratch_dir)
}
if (!dir.exists(cosmicv3_dbs_scratch)) {
  dir.create(cosmicv3_dbs_scratch)
}
```

## Read in data

```{r}
# Declare file path for consensus file
consensus_file <- file.path(data_dir, "snv-consensus-plus-hotspots.maf.tsv.gz")
```

Read in the consensus MAF file. 

```{r}
# Read in the file
maf <- data.table::fread(consensus_file, data.table = FALSE)
```

### Read in color palettes

Read in the histology colors and labels. 

```{r}
histology_label_mapping <- readr::read_tsv(
  file.path(figures_dir, "palettes", "histology_label_color_table.tsv")
) %>% 
  # Select just the columns we will need for plotting
  dplyr::select(Kids_First_Biospecimen_ID, display_group, display_order, hex_codes) %>% 
  # Reorder display_group based on display_order
  dplyr::mutate(display_group = forcats::fct_reorder(display_group, display_order))
```

Set up gradient color palette for the bubble matrix plots. 

```{r}
gradient_col_palette <- readr::read_tsv(
  file.path(figures_dir, "palettes", "gradient_color_palette.tsv")
)

# Won't need NA color this time. 
gradient_col_palette <- gradient_col_palette %>%
  dplyr::filter(color_names != "na_color")
```

Read in the metadata and set it up with the color palette. 

```{r}
metadata_df <- readr::read_tsv(file.path(data_dir, "histologies.tsv"), guess_max = 10000) %>% 
  dplyr::select("Kids_First_Biospecimen_ID", "experimental_strategy") %>%
  dplyr::inner_join(histology_label_mapping, by = "Kids_First_Biospecimen_ID") %>% 
  dplyr::rename(Tumor_Sample_Barcode = "Kids_First_Biospecimen_ID") 
```

Set up BED region files for TMB calculations

```{r}
region_sizes <- readr::read_tsv(file.path(data_dir, "snv-mutation-tmb-all.tsv")) %>%
  dplyr::select(Tumor_Sample_Barcode, region_size)
```

## Set up data

Determine how many mutations we have per sample.

```{r}
mut_per_sample <- maf %>%
  dplyr::group_by(Tumor_Sample_Barcode) %>%
  dplyr::tally() %>%
  dplyr::arrange(n)
summary(mut_per_sample$n)
```

Make mutation data into `deconstructSigs` input format.

```{r}
sigs_input <- mut.to.sigs.input(
  mut.ref = maf,
  sample.id = "Tumor_Sample_Barcode",
  chr = "Chromosome",
  pos = "Start_Position",
  ref = "Reference_Allele",
  alt = "Allele",
  sig.type = 'DBS',
  dbs_table = dbs_possible,
  bsg = BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38
)
```

Add total mutations per sample. 

```{r}
total_muts <- apply(sigs_input, 1, sum)
```


## Determine DBS Signatures for COSMIC

Get list of tumor sample ids. 

```{r}
tumor_sample_ids <- maf %>%
  dplyr::filter(Tumor_Sample_Barcode %in% rownames(sigs_input)) %>%
  dplyr::distinct(Tumor_Sample_Barcode) %>%
  dplyr::pull(Tumor_Sample_Barcode)
```

Get [COSMIC DBS v3.3 signatures](https://cancer.sanger.ac.uk/cosmic) for each sample. 
This step will take some time. 

```{r}
signatures.cosmic.dbs <- read_tsv(file.path(input_dir, 'COSMIC_v3.3_DBS_GRCh38.txt')) %>%
  column_to_rownames('Type') %>%
  t %>%
  as.data.frame()

sample_dbs_sigs_cosmic <- lapply(tumor_sample_ids, function(sample_id) {
  # Determine the signatures contributing to the sample
  whichSignatures(
    tumor.ref = sigs_input,
    signatures.ref = signatures.cosmic.dbs,
    sample.id = sample_id,
    contexts.needed = TRUE
  )
})

# Bring along the names
names(sample_dbs_sigs_cosmic) <- tumor_sample_ids
# 

# Create matrix of COSMIC signature weights
cosmic_dbs_weights <- lapply(sample_dbs_sigs_cosmic, "[[", "weights")
cosmic_dbs_wide <- do.call(dplyr::bind_rows, cosmic_dbs_weights) %>%
  add_column('Kids_First_Biospecimen_ID' = unlist(lapply(cosmic_dbs_weights, rownames)), .before = 1) %>%
  tibble::as_tibble() %>%
  readr::write_tsv(file.path(results_dir, 'cosmicv3.3_DBS_signature_exposure_matrix.tsv')) 
```


Calculate COSMIC DBS signature mutations per Mb for each sample

```{r}
cosmic_dbs_sigs_df <- calc_mut_per_sig(
  sample_dbs_sigs_cosmic,
  muts_per_sample = total_muts,
  region_size = region_sizes,
  metadata = metadata_df
)  %>%
  dplyr::filter(grepl("DBS", signature))

# Write this to a file but drop the color column
cosmic_dbs_sigs_df %>% 
  dplyr::select(-hex_codes) %>% 
  readr::write_tsv(file.path(results_dir, "cosmicv3.3_dbs_signatures_results.tsv"))

# Print out a preview
cosmic_dbs_sigs_df
```


## Mutation signature bubble matrix by histology groups

```{r}
bubble_matrix_plot(cosmic_dbs_sigs_df, 
                   label = "COSMIC DBS Signatures", 
                   color_palette = gradient_col_palette$hex_codes
)
```

```{r}
ggplot2::ggsave(
  file.path(cosmic_dbs_plots, "bubble_matrix_cosmic_dbs_mutation_sig.png"),
  width = 30, height = 20, units = "cm")
```


Mutation signature grouped bar plots for each histology group

```{r, results = "hide"}
# Make grouped bar plots
lapply(unique(cosmic_dbs_sigs_df$display_group),
  grouped_sig_barplot,
  sig_num_df = cosmic_dbs_sigs_df,
  output_dir = file.path(cosmicv3_dbs_scratch, "signature_grouped_barplots"), 
  label = "cosmicv3_dbs"
)
```
